\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing

### 2012-09-24 ipdps13/ace-ipdps13.tex
IPDPS13: Cleaned up abstract, first couple chapters in intro.

\usepackage[noadjust]{cite}
% IEEE wants to use ],\,[ for this
% but that looks dumb...
\renewcommand{\citepunct}{,\,}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

@@

% *** CITATION PACKAGES ***
%
\usepackage{cite}
\renewcommand{\citemid}{,~}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

@@

\begin{abstract}
In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
Ace consists of a statically-typed language that shares syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage. Unlike standard languages, however, the Ace compiler consists only of a minimal core that delegates control over most aspects of front-end compilation to {\em active libraries} using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}. Users specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with designated methods that the compiler selectively invokes when type checking and translating expressions. Using \ATT, we implement the primitives of the C99 and OpenCL languages as modules within Ace. We then give examples of several first-class parallel abstractions that can be used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers developing language-based programming abstractions.

Ace also supports more conventional forms of structured and unstructured metaprogramming and Ace programs can be compiled and launched from Python directly. We used these features, along with the OpenCL language module, to design a scientific simulation framework that allows users to modularly specify and orchestrate the execution of parameterized families of simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing  evidence that Ace is also useful in practice today.
\end{abstract}

@@

Computer-aided simulation and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}. Most of this software targets desktop hardware, and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. To fully harness the power of this modern hardware, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.

Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For tasks where these interpretted languages are too slow, they generally call into code written in statically-typed, low-level languages like C and Fortran that uses low-level concurrency abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. These low-level languages and abstractions are notoriously error-prone and difficult to use, even for expert programmers. {\color{red} CITATION?}

The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns relating to performance, portability, verifiability and ease-of-use. Unfortunately, professional end-users have not adopted these {\em en masse}, and many are generally skeptical of new approaches from the research community. This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those working to advance the practice of scientific and high-performance computing.

One reason for this is that new features and abstractions tend to come in new languages. This is because they need first-class language and compiler support. Monolithic languages make it difficult to provide this. Thus, we need a mechanism that allows new primitives to be defined and implemented modularly and distributed with libraries.

@@

\begin{figure}
\small{
\begin{verbatim}
import clq
from clq.backends.opencl import 
    OpenCL, get_global_id

@clq.fn(OpenCL)
def map(in, out, fn):
    gid = get_global_id(0)
    out[gid] = fn(in[gid])
\end{verbatim}}
\caption{Data-parallel higher-order map in $\cloquence$. The \texttt{get\_global\_id} function is an OpenCL primitive function.}
\label{map}

\end{figure}

\begin{figure}
\small{\begin{verbatim}
#pragma OPENCL EXTENSION cl_khr_fp64 : enable

__kernel void map_sin_dbl(__global double *in, 
                          __global double *out) 
{
    size_t gid = get_global_id(0);
    out[gid] = sin(in[gid]);
}
\end{verbatim}}
\caption{An OpenCL kernel function that maps the sin function over a vector of double-precision floating-point numbers in the global memory address space.}
\label{mapsinint}
\end{figure}

>>>>

\usepackage{cite}
% IEEE wants to use ],\,[ for this
% but that looks dumb...
\renewcommand{\citepunct}{,\,}

%\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

@@

% *** CITATION PACKAGES ***
%
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

@@

\begin{abstract}
We introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
Ace consists of a statically-typed programming language with user-extensible semantics controlled by a compile-time meta\-language, Python. Users specify new primitive types and their operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (\ATT)}. This mechanism allowed us to express the primitives of the C99 and OpenCL languages as simple modules within Ace. We then give examples of several first-class parallel abstractions that can be defined and used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers and domain experts developing first-class language abstractions, as well as practitioners wishing to use multiple such abstractions within a single program. Active library support in Ace also supports more general forms of type-aware metaprogramming, and Ace functions can be compiled and launched directly from Python, taking its standard numeric data structures as arguments.
Using active library support in Ace, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace can be used, at scale, in practice today.
\end{abstract}

@@

Computer-aided simulation and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}. Most of this software targets desktop hardware and about 20\% of scientists target either local clusters or super\-computers for more numerically-intensive computations as well \cite{hannay2009scientists}. To fully harness the power of this modern hardware, however, these {\em professional end-user developers} \cite{segal2007some} must write parallel programs.

Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-intensive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For tasks where these interpretted languages are too slow, they generally call into code written in a statically-typed, low-level language like C and Fortran and use low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously error-prone and difficult to use, even for expert programmers. {\color{red} CITATION?}

The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns relating to performance, portability, verifiability and ease-of-use. Unfortunately, professional end-users have not commonly adopted these approaches, and many have become generally skeptical of new approaches originating in the research community. This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance the practice of scientific and high-performance computing.

One reason for this gap is because new features and abstractions tend to be distributed within new programming languages. Prominent examples include:
\begin{itemize}
\item UPC, CAF, X10
\item Erlang
\item Cilk
\item Charm++
\end{itemize}

This is because they need first-class language and compiler support but monolithic languages cannot be modularly and portably extended (we discuss the difference between language extensions and compiler extensions in Section XX). Unfortunately, writing code in multiple languages poses a host of challenges. Thus, we need a mechanism that allows new primitives to be defined and implemented modularly and distributed with libraries.

@@

\begin{figure}
\small{
\begin{verbatim}
from ace.OpenCL import OpenCL, get_global_id

@OpenCL.fn
def map(in, out, fn):
    gid = get_global_id(0)
    out[gid] = fn(in[gid])
\end{verbatim}
}
\vspace{-10pt}
\caption{A generic data-parallel map function written using the user-defined Ace OpenCL module.}
\label{map}
\end{figure}
\begin{figure}
\small{
\begin{verbatim}
from figure1 import map
from ace.OpenCL import global_ptr, double, sin

T = global_ptr(double)
map_sin_dbl = map.compile(T, T, sin.cl_type)
\end{verbatim}
}
\vspace{-10pt}
\caption{The generic map function compiled to map the sin function over global arrays of double-precision floating-point numbers. The \texttt{get\_global\_id} and \texttt{sin} functions are OpenCL primitives.}
\label{mapsindblace}
\end{figure}
\begin{figure}
\small{\begin{verbatim}
#pragma OPENCL EXTENSION cl_khr_fp64 : enable

__kernel void map_sin_dbl(
  __global double* in, 
  __global double* out) 
{
    size_t gid;
    gid = get_global_id(0);
    out[gid] = sin(in[gid]);
}\end{verbatim}
}
\vspace{-10pt}
\caption{The OpenCL kernel function generated by running \texttt{acecc figure2.py}, or by executing the  \texttt{print map\_sin\_dbl.code} statement.}
\label{mapsinint}
\vspace{-10pt}
\end{figure}
